---
title: "Chapter 4"
subtitle: "Geocentric Models"
format: 
  html: 
    theme: cosmo
editor_options: 
  chunk_output_type: console
---

```{r}
library(tidyverse) # ggplot, dplyr, and friends
library(parameters) # Show model results as nice tables
library(brms) # interface to Stan
library(tidybayes) # Manipulate Stan results in tidy ways
library(ggdist) # Fancy distribution plots
library(patchwork) # Combine ggplot plots
library(rcartocolor) # Color palettes from CARTOColors
library(rstan)
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())
# Custom ggplot theme which i stole from https://www.andrewheiss.com/blog/2023/08/12/conjoint-multilevel-multinomial-guide/#part-1-candy-single-question-basic-multinomial-logit

theme_nice <- function() {
  theme_minimal(base_family = "Clear Sans") +
    theme(
      panel.grid.minor = element_blank(),
      plot.title = element_text(family = "Clear Sans", face = "bold"),
      axis.title.x = element_text(hjust = 0),
      axis.title.y = element_text(hjust = 1),
      strip.text = element_text(
        family = "Clear Sans", face = "bold",
        size = rel(0.75), hjust = 0
      ),
      plot.title.position = "plot",
      strip.background = element_rect(fill = "grey90", color = NA)
    )
}

theme_set(theme_nice())

clrs <- carto_pal(name = "Prism")
```


Linear regression as a baseline thing to learn.
gaussian

Everything that adds together fluctu<ation will sooner or later become gaussian.
Does not have to be purely additive

```{r}
pos <- replicate(1000, sum(runif(16, -1, 1)))
hist(pos)
rethinking::dens(pos)
```

$$
W \sim Binomial(N,p) \\
p \sim Uniform(0,1)
$$

## time to model

```{r}
data(Howell1, package = "rethinking")
d <- Howell1

glimpse(d)

p1 <-
  d |>
  ggplot(aes(x = weight, y = height)) +
  geom_point(col = clrs[9]) +
  geom_smooth(method = "loess", se = F, col = clrs[3]) +
  labs(
    x = "Weight",
    y = "Height",
    title = "Relationship between weight and height \nfor the whole data",
    subtitle = "Clearly non-linear"
  )

d_2 <- d |> filter(age >= 18)

p2 <-
  d_2 |>
  ggplot(aes(x = weight, y = height)) +
  geom_smooth(method = "lm", se = F, col = clrs[3]) +
  geom_point(col = clrs[1]) +
  labs(
    x = "Weight",
    y = "Height",
    title = "Relationship between weight and height \nfor adults",
    subtitle = "More linear"
  )

p1 + p2

ggplot(d_2, mapping = aes(height)) +
  geom_density() +
  labs(title = "Yeah, looks gaussian to me")
```

$$
h_i \sim N(\mu, \sigma) \\
\mu \sim N(178,20) \\
\sigma \sim Uniform(0,50)
$$
plot those priors

```{r}
p1 <-
  tibble(x = seq(100, 250, length.out = 1000)) |>
  mutate(y = dnorm(x, mean = 178, sd = 20)) |>
  # Plotting
  ggplot(aes(x = x, y = y)) +
  geom_line() +
  labs(
    title = bquote("Prior for " ~ mu),
    x = expression(mu),
    y = "Density"
  )
p2 <-
  tibble(x = seq(-10, 60, length.out = 1000)) |>
  mutate(y = dunif(x, 0, 50)) |>
  # Plotting
  ggplot(aes(x = x, y = y)) +
  geom_line() +
  labs(
    title = bquote("Prior for " ~ sigma),
    x = expression(sigma),
    y = "Density"
  )
p1 + p2
```

Prior predictive check:

```{r}
p3 <-
  tibble(
    mu = rnorm(1e4, 178, 20),
    sigma = runif(1e4, 0, 50),
    prior = rnorm(1e4, mu, sigma)
  ) |>
  ggplot(aes(prior)) +
  geom_density(col = "darkorange4") +
  labs(x = NULL, y = NULL, title = "h ~ dnorm(mu, sigma)")

p4 <-
  tibble(
    mu = rnorm(1e4, 178, 100),
    sigma = runif(1e4, 0, 50),
    prior = rnorm(1e4, mu, sigma)
  ) |>
  ggplot(aes(prior)) +
  geom_density(col = "darkorange4") +
  geom_vline(xintercept = 0, lty = "dashed") +
  geom_vline(xintercept = 272, alpha = .9) +
  scale_x_continuous(breaks = c(-128, 0, 178, 484)) +
  labs(x = NULL, y = NULL, title = "h ~ dnorm(mu, sigma)\nmu ~ dnorm(178,100)")

(p1 + p2) / (p3 + p4)
```

```{r}
b1 <- brm(
  data = d_2,
  family = gaussian,
  height ~ 1,
  prior = c(
    prior(normal(178, 20), class = Intercept),
    prior(uniform(0, 50), class = sigma, ub = 50)
  ),
  iter = 2000, warmup = 1000, chains = 4,
  seed = 4
)

b1 |>
  spread_draws(b_Intercept, sigma) |>
  median_qi(.width = .89) |>
  select(b_Intercept, sigma, .width) |>
  knitr::kable(digits = 3)

b2 <- brm(
  data = d_2,
  family = gaussian,
  height ~ 1,
  prior = c(
    prior(normal(178, .1), class = Intercept),
    prior(uniform(0, 50), class = sigma, ub = 50)
  ),
  iter = 2000, warmup = 1000, chains = 4,
  seed = 4
)

b2 |>
  spread_draws(b_Intercept, sigma) |>
  median_qi(.width = .89) |>
  select(b_Intercept, sigma, .width) |>
  knitr::kable(digits = 2)

b2 |>
  spread_draws(b_Intercept, sigma) |>
  median_qi(condition_mean = b_Intercept + sigma, .width = c(.95)) |>
  ggplot(aes(x = condition_mean)) +
  stat_halfeye()
```

## Linear Prediction



```{r}
d_2 <- d_2 |>
  mutate(weight_c = weight - mean(weight))

b3 <- brm(
  data = d_2,
  family = gaussian,
  height ~ 1 + weight_c,
  prior = c(
    prior(normal(178, 20), class = Intercept),
    prior(lognormal(0, 1), class = b),
    prior(uniform(0, 50), class = sigma, ub = 50)
  ),
  iter = 2000, warmup = 1000, chains = 4,
  seed = 4
)
```

```{r}
plot(b3)

model_parameters(b3,
  ci = .89
)

pairs(b3)

labels <-
  c(-10, 0, 10) + mean(d_2$weight) |>
  round(digits = 0)

d_2 |>
  ggplot(aes(weight_c, height)) +
  geom_abline(
    intercept = fixef(b3)[1],
    slope = fixef(b3)[2],
    col = "firebrick", size = .5
  ) +
  geom_point(shape = 1, size = 2, color = "royalblue") +
  scale_x_continuous("weight",
    breaks = c(-10, 0, 10),
    labels = labels
  ) +
  labs(
    title = "Our cute little model",
    x = "Weight (centred)",
    y = "Height"
  )
```

## 4.5 Curves from lines

```{r}
d <- Howell1

p1 <-
  d |>
  mutate(weight_s = (weight - mean(weight)) / sd(weight)) |>
  ggplot(aes(x = weight_s, y = height)) +
  geom_smooth(
    formula = y ~ x, se = T, col = clrs[5], method = "lm",
    fill = "grey83"
  ) +
  geom_point(
    shape = 1, size = 3, alpha = 1 / 3,
    col = "navy"
  ) +
  labs(
    y = "Height",
    subtitle = "Linear"
  )
```

Polynomial regression: Variablen hoch irgendwas setzen und als weiteren Prädiktor verwenden

$$
\mu_i = \alpha + \beta_1x_i + \beta_2x_i^2
$$
$$
h_i \sim N(\mu_i,\sigma) \\
\mu_i = \alpha + \beta_1x_i + \beta_2x_i^2 \\
\alpha \sim N(178,20) \\
\beta_1 \sim LogNormal(0,1) \\
\beta_2 \sim N(0,1) \\
\sigma \sim U(0,50)
$$


```{r}
d <-
  d |>
  mutate(
    weight_s = (weight - mean(weight)) / sd(weight),
    weight_s2 = weight_s^2
  )

b5 <- brm(
  data = d,
  family = gaussian,
  height ~ 1 + weight_s + weight_s2,
  prior = c(
    prior(normal(178, 20), class = Intercept),
    prior(lognormal(0, 1), class = b, coef = "weight_s"),
    prior(normal(0, 1), class = b, coef = "weight_s2"),
    prior(uniform(0, 50), class = sigma, ub = 50)
  ),
  iter = 2000, warmup = 1000, chains = 4,
  seed = 4
)

parameters::model_parameters(b5)

weight_seq <-
  tibble(weight_s = seq(from = -2.5, to = 2.5, length.out = 29 + 1)) |>
  mutate(weight_s2 = weight_s^2)

fit <-
  fitted(b5,
    newdata = weight_seq
  ) |>
  data.frame() |>
  bind_cols(weight_seq)

pred <- predict(b5,
  newdata = weight_seq
) |>
  data.frame() |>
  bind_cols(weight_seq)

p2 <-
  ggplot(
    data = d,
    aes(x = weight_s)
  ) +
  geom_ribbon(
    data = pred,
    aes(ymin = Q2.5, ymax = Q97.5),
    fill = "grey83"
  ) +
  geom_smooth(
    data = fit,
    aes(y = Estimate, ymin = Q2.5, ymax = Q97.5),
    stat = "identity",
    fill = "grey70", color = "black", alpha = 1, linewidth = 1 / 2
  ) +
  geom_point(aes(y = height),
    color = "navy", shape = 1, size = 3, alpha = 1 / 3
  ) +
  labs(
    subtitle = "Quadratic",
    y = "Height"
  ) +
  coord_cartesian(
    xlim = range(d$weight_s),
    ylim = range(d$height)
  )

p2
```


$$
h_i \sim N(\mu_i,\sigma) \\
\mu_i = \alpha + \beta_1x_i + \beta_2x_i^2 + \beta_3x_i^3\\
\alpha \sim N(178,20) \\
\beta_1 \sim LogNormal(0,1) \\
\beta_2 \sim N(0,1) \\
\beta_3 \sim N(0,1) \\
\sigma \sim U(0,50)
$$

```{r}
d <-
  d |>
  mutate(weight_s3 = weight_s^3)


b6 <- brm(
  data = d,
  family = gaussian,
  height ~ 1 + weight_s + weight_s2 + weight_s3,
  prior = c(
    prior(normal(178, 20), class = Intercept),
    prior(lognormal(0, 1), class = b, coef = "weight_s"),
    prior(normal(0, 1), class = b, coef = "weight_s2"),
    prior(normal(0, 1), class = b, coef = "weight_s3"),
    prior(uniform(0, 50), class = sigma, ub = 50)
  ),
  iter = 2000, warmup = 1000, chains = 4,
  seed = 4
)

parameters::model_parameters(b6)

weight_seq <-
  weight_seq |>
  mutate(weight_s3 = weight_s^3)

fit <-
  fitted(b6,
    newdata = weight_seq
  ) |>
  data.frame() |>
  bind_cols(weight_seq)

pred <- predict(b6,
  newdata = weight_seq
) |>
  data.frame() |>
  bind_cols(weight_seq)

p3 <-
  ggplot(
    data = d,
    aes(x = weight_s)
  ) +
  geom_ribbon(
    data = pred,
    aes(ymin = Q2.5, ymax = Q97.5),
    fill = "grey83"
  ) +
  geom_smooth(
    data = fit,
    aes(y = Estimate, ymin = Q2.5, ymax = Q97.5),
    stat = "identity",
    fill = "grey70", color = "black", alpha = 1, linewidth = 1 / 2
  ) +
  geom_point(aes(y = height),
    color = "navy", shape = 1, size = 3, alpha = 1 / 3
  ) +
  labs(
    subtitle = "Kubic",
    y = "Height"
  ) +
  coord_cartesian(
    xlim = range(d$weight_s),
    ylim = range(d$height)
  )

(p1 + p2 + p3)

b4 <-
  brm(
    data = d,
    family = gaussian,
    height ~ 1 + weight_s,
    prior = c(
      prior(normal(178, 20), class = Intercept),
      prior(lognormal(0, 1), class = b, coef = "weight_s"),
      prior(uniform(0, 50), class = sigma, ub = 50)
    ),
    iter = 2000, warmup = 1000, chains = 4, cores = 4,
    seed = 4
  )

performance::compare_performance(b4, b5, b6)
```

### splines

yeah no idea how to do this with brms

```{r}
data(cherry_blossoms, package = "rethinking")
d <- cherry_blossoms
rethinking::precis(d)

d |>
  ggplot(aes(year, doy)) +
  geom_point(shape = 1, alpha = .3) +
  geom_hline(yintercept = 100, lty = "dashed") +
  geom_smooth(se = F)

d2 <- d[complete.cases(d$doy), ]

num_knots <- 15

knots_list <- quantile(d2$year, probs = seq(0, 1, length.out = num_knots))

library(splines)
B <- bs(d2$year,
  knots = knots_list[-c(1, num_knots)],
  degree = 3, intercept = T
)

plot(NULL,
  xlim = range(d2$year),
  ylim = c(0, 1),
  xlab = "Year",
  ylab = "Basis"
)
for (i in 1:ncol(B)) lines(d2$year, B[, i])

ms <- rethinking::quap(
  alist(
    D ~ dnorm(mu, sigma),
    mu <- a + B %*% w,
    a ~ dnorm(100, 10),
    w ~ dnorm(0, 10),
    sigma ~ dexp(1)
  ),
  data = list(D = d2$doy, B = B),
  start = list(w = rep(0, ncol(B)))
)
post <- rethinking::extract.samples(ms)

w <- apply(post$w, 2, mean)

plot(
  NULL,
  xlim = range(d2$year),
  ylim = c(-6, 6),
  xlab = "Year",
  ylab = "Basis * Weight"
)

for (i in ncol(B)) {
  lines(d2$year, w[i] * B[, i])
}

mu <- link(ms)
mu_PI <- apply(mu, 2, PI, 0.97)
plot(d2$year, d2$doy, col = col.alpha(rangi2, 0.3), pch = 16)
shade(mu_PI, d2$year, col = col.alpha("black", 0.5))
```

# Exercise

## 4E1
The first line is the likelihood

## 4E2
Two, $\mu$ and $\sigma$

## 4E3

$$
Pr(\mu,\sigma|y) = \frac{Pr(y|\mu,\sigma)Pr(\mu)Pr(\sigma)}{\int\int Pr(y|\mu,\sigma)Pr(\mu)Pr(\sigma)d\mu d\sigma}
$$

## 4E4

The second line

## 4E5

3, \alpha, \beta, \sigma

## 4M1

```{r}
tibble(
  n = 100000,
  mu = rnorm(n, 0, 10),
  sigma = rexp(n, 1)
) |>
  mutate(y = rnorm(n, mu, sigma)) |>
  # Plotting
  ggplot(aes(y)) +
  geom_density() +
  labs(
    title = bquote("Prior for"),
    x = expression(mu),
    y = "Density"
  )
```

## 4M2
```{r}
#| eval: false
brm(
  data = d,
  family = gaussian,
  formula = y ~ 1,
  prior = c(
    prior(normal(0, 10), class = Intercept),
    prior(exponential(1), class = sigma)
  ),
  iter = 2000, warmup = 1000, chains = 4, cores = 4,
  seed = 4
)
```

## 4M3

y ~ dnorm( mu , sigma ),
mu <- a + b*x,
a ~ dnorm( 0 , 10 ),
b ~ dunif( 0 , 1 ),
sigma ~ dexp( 1 )


$$
y_i \sim N(\mu,\sigma) \\
\mu_i = \alpha + \beta x_i \\
\alpha \sim N(0,10) \\
\beta \sim Uniform(0,1) \\
\sigma \sim Exponential (1)
$$

## 4M5

$$
Height \sim Normal(\mu,\sigma) \\
\mu_i = \alpha + \beta x_i \\
\alpha \sim Normal(100,20) \\
\beta \sim Lognormal(1,0.5) \\
\sigma \sim Exponential(1)
$$

```{r}
n <- 50

# Create a data frame
data <- tibble(
  i = rep(1:n, each = 3),
  s = rep(rexp(n, 1), each = 3),
  b = rep(rlnorm(n, 1, 0.5), each = 3),
  a = rep(rnorm(n, 100, 20), each = 3),
  y = rep(1:3, n)
) |>
  mutate(ybar = mean(y)) |>
  rowwise() |>
  mutate(h = rnorm(1, mean = a + b * (y - ybar), sd = s)) |>
  ungroup()

# Reshaping data for plotting
long_data <- data |>
  select(i, year = y, h)

# Plot using ggplot2
ggplot(long_data, aes(x = year, y = h, group = i)) +
  geom_line() +
  labs(x = "Year", y = "Height (cm)") +
  scale_x_continuous(breaks = 1:3) +
  theme(
    panel.grid = element_blank()
  )
```

## 4M6
Put a uniform prior on the variance, like $\sigma \sim Uniform(0,8)$, $8$ because we use the sd not the var here and ofc $\sqrt{64} = 8$

## 4M7

```{r}
b_no_c <- brm(
  data = d_2,
  family = gaussian,
  height ~ 1 + weight,
  prior = c(
    prior(normal(178, 20), class = Intercept),
    prior(lognormal(0, 1), class = b),
    prior(uniform(0, 50), class = sigma, ub = 50)
  ),
  iter = 2000, warmup = 1000, chains = 4,
  seed = 4
)

post <- as_draws_df(b3)
# variances
select(post, b_Intercept:sigma)  |> 
  cov() |> 
  diag()

post <- as_draws_df(b_no_c)
# variances
select(post, b_Intercept:sigma)  |> 
  cov() |> 
  diag()

post_pred_3 <- posterior_predict(b3)

y_pred <- apply(post_pred_3, 2, mean)

p1 <- 
ggplot(d_2, aes(x = weight, y = height)) +
  geom_point(shape = 1, alpha = .33) +
  geom_line(aes(y = y_pred), color = "steelblue") +
  labs(title = "Weight centred")


post_pred_no_c <- posterior_predict(b_no_c)

y_pred_no_c <- apply(post_pred_no_c, 2, mean)

p2 <- ggplot(d_2, aes(x = weight, y = height)) +
  geom_point(shape = 1, alpha = .33) +
  geom_line(aes(y = y_pred_no_c), color = "firebrick") +
  labs(title = "Weight not centred")

p1 + p2
```

Slope is different, which makes sense, yet the posterior prediction is basically the same
